Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods

### 2023-07-28 paper/main.tex

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle
\begin{abstract}
  This paper introduces a novel computational framework for solving
alternating current optimal power flow (ACOPF) problems using
graphics processing units (GPUs). While GPUs have demonstrated
remarkable performance in various computing domains, their application
in AC OPF has been limited due to challenges associated with porting
sparse automatic differentiation (AD) and sparse linear solver
routines to GPUs. We aim to address these issues with two key
strategies. First, we utilize a single-instruction, multiple-data
(SIMD) abstraction of nonlinear programs (NLP). This approach enables
the specification of model equations while preserving their
parallelizable structure, and in turn, facilitates the implementation
of AD routines that can exploit such structure. Second, we employ a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy. This technique involves relaxing equality
constraints to inequalities and condensing the Karush-Kuhn-Tucker
system into a much smaller positive definite system. This strategy
offers the key advantage of being able to factorize the KKT matrix
without numerical pivoting, which in the past has hampered the parallelization of
the IPM algorithm. By combining these two strategies, we can perform the
majority of operations on GPUs while keeping the data residing in the
device memory only. Comprehensive numerical benchmark results showcase
the substantial computational advantage of our approach. Remarkably,
for solving large-scale AC OPF problems to a moderate accuracy, our
implementations—MadNLP.jl and SIMDiff.jl—running on NVIDIA GPUs
achieve an order of magnitude speedup compared to state-of-the-art
tools running on contemporary CPUs.
\end{abstract}

@@

Now, we observe that a significant portion of the system in Fig.
\eqref{fig:very-long-eqn} can be eliminated by exploiting the block
structure, leading to an equivalent system stated in a smaller
space. In particular, the lower-right $4\times 4$ block is always
invertible since the IPM procedure ensures that the iterates stay in
the strict interior of the feasible set. This allows for eliminating
the lower-right 4x4 block, resulting in:
\begin{align}\label{eqn:very-long-reduced}
  &
    \underbrace{
    \begin{bmatrix}
      W  + \Sigma_x + \delta_w I & & A^{\top} \\
      & \Sigma_s + \delta_w I & -I\\
      A& -I & -\delta_c I\\
    \end{bmatrix}
  }_{M_{\text{red}}}
  \begin{bmatrix}
    \Delta x \\
    \Delta s \\
    \Delta y \\
  \end{bmatrix}=
  \begin{bmatrix}
    q_x \\
    q_s\\
    q_y\\
  \end{bmatrix},
\end{align}
where:
\begin{align*}
  \Sigma_x&:= Z^\flat_x (X-X^\flat)^{-1}+ Z^\sharp_x (X^\sharp-X)^{-1}\\
  \Sigma_s&:= Z^\flat_s (S-S^\flat)^{-1}+ Z^\sharp_s (S^\sharp-S)^{-1}\\
  q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
  q_s&:= (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
  q_y&:=p_y.
\end{align*}
The bound dual steps can be recovered as follows:
\begin{align}\label{eqn:recover-1}
  \begin{aligned}[t]
    \Delta z^\flat_x &= \left(X-X^\flat\right)^{-1} \left(-Z^\flat_x \Delta x  + p_{z^\flat_x}\right)\\
    \Delta z^\sharp_x &= \left(X^\sharp-X\right)^{-1} \left(Z^\sharp_x \Delta x  + p_{z^\sharp_x}\right)\\
    \Delta z^\flat_s &= \left(S-S^\flat\right)^{-1} \left(-Z^\flat_s \Delta s  + p_{z^\flat_s}\right)\\
    \Delta z^\sharp_s &= \left(S^\sharp-S\right)^{-1} \left(Z^\sharp_s \Delta s  + p_{z^\sharp_s}\right).
  \end{aligned}
\end{align}
Note that the matrices involved in the inversions in \eqref{eqn:recover-1}
are always diagonal, so their computation is cheap.
Also, note that the reduced system in \eqref{eqn:very-long-reduced}
corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
in the original version of the algorithm, we did not introduce the
slack variables, so it does not have the additional structure imposed
by the slack variables.

>>>>

\date{\small
  $^{1}$Mathematics and Computer Science Division, Argonne National Laboratory\\
  $^{2}$Centre Automatique et Syst\`{e}mes, Mines Paris - PSL
}
\begin{document}
\maketitle
\begin{abstract}
  This paper introduces a novel computational framework for solving
alternating current optimal power flow (ACOPF) problems using
graphics processing units (GPUs). While GPUs have demonstrated
remarkable performance in various computing domains, their application
in AC OPF has been limited due to challenges associated with porting
sparse automatic differentiation (AD) and spars e linear solver
routines to GPUs. We aim to address these issues with two key
strategies. First, we utilize a single-instruction, multiple-data
(SIMD) abstraction of nonlinear programs (NLP). This approach enables
the specification of model equations while preserving their
parallelizable structure, and in turn, facilitates the implementation
of AD routines that can exploit such structure. Second, we employ a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy. This technique involves relaxing equality
constraints to inequalities and condensing the Karush-Kuhn-Tucker
system into a much smaller positive definite system. This strategy
offers the key advantage of being able to factorize the KKT matrix
without numerical pivoting, which in the past has hampered the parallelization of
the IPM algorithm. By combining these two strategies, we can perform the
majority of operations on GPUs while keeping the data residing in the
device memory only. Comprehensive numerical benchmark results showcase
the substantial computational advantage of our approach. Remarkably,
for solving large-scale AC OPF problems to a moderate accuracy, our
implementations—MadNLP.jl and SIMDiff.jl—running on NVIDIA GPUs
achieve an order of magnitude speedup compared to state-of-the-art
tools running on contemporary CPUs.
\end{abstract}

@@

Now, we observe that a significant portion of the system in Fig.
\eqref{fig:very-long-eqn} can be eliminated by exploiting the block
structure, leading to an equivalent system stated in a smaller
space. In particular, the lower-right $4\times 4$ block is always
invertible since the IPM procedure ensures that the iterates stay in
the strict interior of the feasible set. This allows for eliminating
the lower-right 4x4 block, resulting in:
\begin{align}\label{eqn:very-long-reduced}
  &
    \underbrace{
    \begin{bmatrix}
      W  + \Sigma_x + \delta_w I & & A^{\top} \\
      & \Sigma_s + \delta_w I & -I\\
      A& -I & -\delta_c I\\
    \end{bmatrix}
  }_{M_{\text{aug}}}
  \begin{bmatrix}
    \Delta x \\
    \Delta s \\
    \Delta y \\
  \end{bmatrix}=
  \begin{bmatrix}
    q_x \\
    q_s\\
    q_y\\
  \end{bmatrix},
\end{align}
where:
\begin{align*}
  \Sigma_x&:= Z^\flat_x (X-X^\flat)^{-1}+ Z^\sharp_x (X^\sharp-X)^{-1}\\
  \Sigma_s&:= Z^\flat_s (S-S^\flat)^{-1}+ Z^\sharp_s (S^\sharp-S)^{-1}\\
  q_x&:=p_x + (X-X^\flat)^{-1} p_{z^\flat_x}-  (X^\sharp-X)^{-1} p_{z^\sharp_x}\\
  q_s&:= (S-S^\flat)^{-1} p_{z^\flat_s}-  (S^\sharp-S)^{-1} p_{z^\sharp_s}\\
  q_y&:=p_y.
\end{align*}
The bound dual steps can be recovered as follows:
\begin{align}\label{eqn:recover-1}
  \begin{aligned}[t]
    \Delta z^\flat_x &= \left(X-X^\flat\right)^{-1} \left(-Z^\flat_x \Delta x  + p_{z^\flat_x}\right)\\
    \Delta z^\sharp_x &= \left(X^\sharp-X\right)^{-1} \left(Z^\sharp_x \Delta x  + p_{z^\sharp_x}\right)\\
    \Delta z^\flat_s &= \left(S-S^\flat\right)^{-1} \left(-Z^\flat_s \Delta s  + p_{z^\flat_s}\right)\\
    \Delta z^\sharp_s &= \left(S^\sharp-S\right)^{-1} \left(Z^\sharp_s \Delta s  + p_{z^\sharp_s}\right).
  \end{aligned}
\end{align}
Note that the matrices involved in the inversions in \eqref{eqn:recover-1}
are always diagonal, so their computation is cheap.
Also, note that the augmented system in \eqref{eqn:very-long-reduced}
corresponds to the KKT system in \eqref{eqn:kkt-indefinite}. However,
in the original version of the algorithm \cite{pacaud2023accelerating}, we did not introduce the
slack variables, so it did not have the additional structure imposed
by the slack variables.
